---
title: "Familiality of MTX persistence - post-hoc analysis"
date: "December 2021"
output:
  word_document: default
  html_document: default
---

```{r, echo = FALSE}
### NOTE: File paths are not very stable and it is likely that files have been moved since last usage...
```

### INTRODUCTION

In this document I extend our primary results from the study of the familiality of MTX persistence.
I do this through a post-hoc analysis assessing how socioeconomy and geography interplays with persistence.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(dplyr)
library(haven)
library(ggplot2)
```

```{r, echo = FALSE}
df_edu <- read_sas("H:/Projekt/Till_Anton_socio_plus_acpa/cohort.sas7bdat")
df_edu <- df_edu %>%
  select(pid, EDU, p1yr, p3yr, p1yr_rel, p3yr_rel) %>%    #Extract only relevant columns
  distinct() %>%    #Remove any duplicated line
  mutate(EDU = ifelse(EDU %in% c("0", "1", "2"), "<9", EDU)) %>%
  mutate(EDU = ifelse(EDU == "3", "9-12", EDU)) %>%
  mutate(EDU = ifelse(EDU %in% c("4", "5", "6"), ">12", EDU)) %>%
  mutate(EDU = ifelse(EDU == "<9", 0, ifelse(EDU == "9-12", 1, 2)))

df_geo <- read_sas("H:/Projekt/Familiarity of mono-MTX/Data Extraction/Misc/clin_reg.sas7bdat")
```

### SOCIOECONOMY

As a proxy for socioeconomy, I use *educational attainment*, here classified as a categorical variable with three levels, each based on years of education: $< 9$; $9-12$; $>12$.
Data on educational attainment was obtained via the LISA register.

When observing the proportion of persistent patients across attained educational level, a clear trend is noticed where the proportion persistent patients decreases as educational level increases (Figure 1).
A similar relationship is observable for persistence at three years (Appendix, Figure A1).

```{r, echo = FALSE, message = FALSE}
df_edu %>%
  group_by(EDU, p1yr) %>% summarise(Np1yr = n()) %>%
  left_join(df_edu %>% group_by(EDU) %>% summarise(N_tot = n()), by = "EDU") %>%
  mutate(FREQ = Np1yr / N_tot) %>%
  ggplot(aes(x = factor(EDU), y = FREQ, fill = factor(p1yr))) + 
  geom_col() +
  geom_hline(yintercept = 0.66, alpha = 0.5, linetype = "dashed", col = "Grey 3") + 
  xlab("Educational level") + ylab("Proportion") + 
#  ggtitle("Proportion persistent patients across attained \neducational level; \nDashed line represents population #average") +
  scale_x_discrete(labels = c("<9", "9-12", ">12")) +
  scale_fill_discrete(labels = c("Non-persistent", "Persistent"),
                      name = "Persistence at \none year") +
  theme(legend.position = "none")
```

However, what is important for us here, is that educational attainment (socioeconomy) is not a confounder.
We can assess this by inspecting how it affects the proportion of concordant (i.e., pairs where *both* the patient and their relative had the same persistence status) and discordant pairs.

A similar relationship is discernible when looking at the pairwise concordance of persistence status (as opposed to persistence status of the index patients) (Figure 2).
However, the deviation from the population average is less dramatic, and the effect is seemingly attenuated.
A similar, yet further attenuated effect, is observable for persistence at three years (Appendix, Figure A2).
In sum, I find that these results indicate two things.
Firstly, socioeconomy *has* an association with persistence status, in the sense that eductional attainment is negatively correlated with persistence.
Secondly, this effect remains when looking at the pairwise concordance of persistence status though it is increasingly attenuated with time, making it unlikely to be driving our results.

```{r, echo = FALSE, message = FALSE}
df_edu %>%
  mutate(concordant_p1yr = ifelse(p1yr == p1yr_rel, 1, 0)) %>%
  group_by(EDU, concordant_p1yr) %>% summarise(Np1yr = n()) %>%
  left_join(df_edu %>% group_by(EDU) %>% summarise(N_tot = n()), by = "EDU") %>%
  mutate(FREQ = Np1yr / N_tot) %>%
  ggplot(aes(x = factor(EDU), y = FREQ, fill = factor(concordant_p1yr))) +
  geom_col() +
  geom_hline(yintercept = 0.570, alpha = 0.5, linetype = "dashed", col = "Grey 3") +
  xlab("Educational level") + ylab("Proportion") +
#  ggtitle("Proportion pairs concordant for persistence status \nacross educational level; \nDashed line represent #population average") +
  scale_x_discrete(labels = c("<9", "9-12", ">12")) +
  scale_fill_discrete(labels = c("Discordant", "Concordant"),
                      name = "Persistence status at \none year") +
  theme(legend.position = "none")
```

### GEOGRAPHY

During my Tuesday morning presentation (16/11), a good point was raised suggesting that geography may influence persistence and thus our results.
It is possible that certain clinics are more (or less) prone to introduce non-MTX DMARDs or to stop treatment with MTX entirely (which would lead to a patient being considered non-persistent) than other clinics are, thus introducing a geographical bias where such clinics have more persistent (or non-persistent) cases than others.

Indeed, the proportion of persistent individuals varies greatly from clinic to clinic (Figure 3a).
However, the more extreme clinics that deviate from the population average are predominantly clinics with few patients, such observations possibly being due to pure chance.
In fact, when removing all clinics with less than five patients in total, results are far more stable (Figure 3b).
Results are similar for persistence at three years (Appendix, Figure 4).

```{r, echo = FALSE, warning = FALSE, message = FALSE}
df_geo %>%
  group_by(tillhor, persist_1yr) %>% summarise(Np1yr = n()) %>%
  left_join(df_geo %>% group_by(tillhor) %>% summarise(N_tot = n()), by = "tillhor") %>%
  mutate(FREQ = Np1yr / N_tot) %>%
  ggplot(aes(x = factor(tillhor), y = FREQ, fill = factor(persist_1yr))) + 
  geom_col() +
  geom_hline(yintercept = 0.66, alpha = 0.5, linetype = "dashed", col = "Grey 3") +
  coord_flip() + xlab("") + ylab("Proportion, scaled to clinic size") +
#  ggtitle("Proportion persistent patients per clinic \nDashed line represents the general proportion") +
  scale_fill_discrete(labels = c("Non-persistent", "Persistent"), 
                      name = "Peristence \nat one year") +
  theme(legend.position = "none")
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
df_geo %>%
  group_by(tillhor, persist_1yr) %>% summarise(Np1yr = n()) %>%
  left_join(df_geo %>% group_by(tillhor) %>% summarise(N_tot = n()), by = "tillhor") %>%
  filter(N_tot > 5) %>%    #Filter based on cutoff
  mutate(FREQ = Np1yr / N_tot) %>%
  ggplot(aes(x = factor(tillhor), y = FREQ, fill = factor(persist_1yr))) + 
  geom_col() +
  geom_hline(yintercept = 0.66, alpha = 0.5, linetype = "dashed", col = "Grey 3") +
  coord_flip() + xlab("") + ylab("Proportion, scaled to clinic size") +
#  ggtitle("Proportion persistent patients per clinic \nDashed line represents the general proportion") +
  scale_fill_discrete(labels = c("Non-persistent", "Persistent"), 
                      name = "Peristence \nat one year") +
  theme(legend.position = "none")
```

This could affect our results if concordant pairs (i.e., pairs where *both* the patient and their relative had the same persistence status) were more often visiting the same clinic.
However, when comparing the proportion of pairs concordant for geography, both marginal as well as stratified on concordance for persistence status, results are near-identical (Table 1).
In sum, I find that these results indicate that it is unlikely that geography has a relevant effect on our results.

```{r, echo = FALSE, results = FALSE}
df_geo %>% filter(persist_1yr == rel_persist_1yr) %>% nrow()    #Number of concordant pairs 1 year
df_geo %>% filter(persist_3yr == rel_persist_3yr) %>% nrow()    #Number of concordant pairs 3 years

df_geo %>% filter(persist_1yr == rel_persist_1yr) %>%
  filter(tillhor == tillhor_rel) %>% nrow()    #Number of concordant pairs concordant for clinic 1 year
df_geo %>% filter(persist_3yr == rel_persist_3yr) %>%
  filter(tillhor == tillhor_rel) %>% nrow()    #Number of concordant pairs concordant for clinic 3 years

df_geo %>% filter(persist_1yr == rel_persist_1yr) %>%
  filter(lan == lan_rel) %>% nrow()    #Number of concordant pairs concordant for lan 1 year
df_geo %>% filter(persist_3yr == rel_persist_3yr) %>%
  filter(lan == lan_rel) %>% nrow()    #Number of concordant pairs concordant for lan 3 years

df_geo %>% filter(persist_1yr == rel_persist_1yr) %>%
  filter(region == region_rel) %>% nrow()    #Number of concordant pairs concordant for region 1 year
df_geo %>% filter(persist_3yr == rel_persist_3yr) %>%
  filter(region == region_rel) %>% nrow()    #Number of concordant pairs concordant for region 1 year
```

### APPENDIX

```{r, echo = FALSE, message = FALSE}
df_edu %>%
  group_by(EDU, p3yr) %>% summarise(Np3yr = n()) %>%
  left_join(df_edu %>% group_by(EDU) %>% summarise(N_tot = n()), by = "EDU") %>%
  mutate(FREQ = Np3yr / N_tot) %>%
  ggplot(aes(x = factor(EDU), y = FREQ, fill = factor(p3yr))) + 
  geom_col() +
  geom_hline(yintercept = 0.50, alpha = 0.5, linetype = "dashed", col = "Grey 3") + 
  xlab("Educational level") + ylab("Proportion") + 
#  ggtitle("Proportion persistent patients across attained educational level \nDashed line represents population #average") +
  scale_x_discrete(labels = c("<9", "9-12", ">12")) +
  scale_fill_discrete(labels = c("Non-persistent", "Persistent"),
                      name = "Persistence at \nthree years") +
  theme(legend.position = "none")
```

```{r, echo = FALSE, message = FALSE}
df_edu %>%
  mutate(concordant_p3yr = ifelse(p3yr == p3yr_rel, 1, 0)) %>%
  group_by(EDU, concordant_p3yr) %>% summarise(Np3yr = n()) %>%
  left_join(df_edu %>% group_by(EDU) %>% summarise(N_tot = n()), by = "EDU") %>%
  mutate(FREQ = Np3yr / N_tot) %>%
  ggplot(aes(x = factor(EDU), y = FREQ, fill = factor(concordant_p3yr))) +
  geom_col() +
  geom_hline(yintercept = 0.595, alpha = 0.5, linetype = "dashed", col = "Grey 3") +
  xlab("Educational level") + ylab("Proportion") +
#  ggtitle("Proportion pairs concordant for persistence status across educational level \nDashed line represent #population average") +
  scale_x_discrete(labels = c("<9", "9-12", ">12")) +
  scale_fill_discrete(labels = c("Discordant", "Concordant"),
                      name = "Persistence status at \nthree years") +
  theme(legend.position = "none")
```

```{r, echo = FALSE, message = FALSE}
df_geo %>%
  group_by(tillhor, persist_3yr) %>% summarise(Np3yr = n()) %>%
  left_join(df_geo %>% group_by(tillhor) %>% summarise(N_tot = n()), by = "tillhor") %>%
  mutate(FREQ = Np3yr / N_tot) %>%
  ggplot(aes(x = factor(tillhor), y = FREQ, fill = factor(persist_3yr))) + 
  geom_col() +
  geom_hline(yintercept = 0.50, alpha = 0.5, linetype = "dashed", col = "Grey 3") +
  coord_flip() + xlab("") + ylab("Proportion, scaled to clinic size") +
#  ggtitle("Proportion persistent patients per clinic \nDashed line represents the general proportion") +
  scale_fill_discrete(labels = c("Non-persistent", "Persistent"), 
                      name = "Peristence \nat three years") +
  theme(legend.position = "none")
```

```{r, echo = FALSE, message = FALSE}
df_geo %>%
  group_by(tillhor, persist_3yr) %>% summarise(Np3yr = n()) %>%
  left_join(df_geo %>% group_by(tillhor) %>% summarise(N_tot = n()), by = "tillhor") %>%
  filter(N_tot > 5) %>%    #Filter based on cutoff
  mutate(FREQ = Np3yr / N_tot) %>%
  ggplot(aes(x = factor(tillhor), y = FREQ, fill = factor(persist_3yr))) + 
  geom_col() +
  geom_hline(yintercept = 0.50, alpha = 0.5, linetype = "dashed", col = "Grey 3") +
  coord_flip() + xlab("") + ylab("Proportion, scaled to clinic size") +
#  ggtitle("Proportion persistent patients per clinic \nDashed line represents the general proportion") +
  scale_fill_discrete(labels = c("Non-persistent", "Persistent"), 
                      name = "Peristence \nat three years") +
  theme(legend.position = "none")
```

